Continous time Markov chain models; \\ extension, simulation and visualisation

### 2014-07-08 Chapter5.tex
workin gon introduction

First computationally trackable models of codon substitution were independently proposed by \cite{Muse1994} and \cite{Goldman1994} and published in the same issue of Molecular Biology and Evolution (MBE).
For codon models the non-stop codon triplet $n_{1}n_{2}n_{3}$ is considered as a smallest unit of evoluton.
There are $4^3$ possible triplets minus three stop-codons, resulting in a state space size of $61$ codons.
The standard assuption of independence is used, i.e. the substitutions at these three codon positions occur independently, thus only a single change per triplet can occur at a given time.
There are several advantages in using codon models over nucleotide based substitution models.
Not all DNA positions evolve at the same rate, with non-synonymous substitutions occuring more frequently the synonymous substitutions.
Although this problem can be mitigated to some extent by using codon-positioned nucleotide substitution models, the fast evolving positions and the state space limited to 4 character alphabet leads to biased estimates of long evolutionary distances, as portrayed in the previous chapter (\ref{sub:deepRoot}).

@@

Different flavours of the GY94 model differ in the composition of the equilibrium codon frequency parameter $\pi_{j}$.
One approach is to model the codon frequencies as each having the same long-time frequency of apperaring and such a model is reffered to as the F-equal.
In the GY94-F$1\times4$ model, each codon position has

codon frequencies are expected from the four nucleotide frequencies, giving three free parameters, as they need to some to one.

a separate set of nucleotide propensity parameters to each of the three codon positions.

For the GY94-F$3\times4$ model frequencies are parametrized according to three sets of nucleotide frequencies for the three codon positions, resulting in nine free parameters.
Finally in the GY94-F61 model every codon triplet has it's own frequncy parameter, with all parameters summing to one, resulting in 60 free parameter that need to be estimated.

>>>>

First computationally trackable models of codon substitution were independently proposed by \cite{Muse1994} and \cite{Goldman1994} and published in the same issue of Molecular Biology and Evolution (MBE).
For codon models the non-stop codon triplet $n_{1}n_{2}n_{3}$ is considered as a smallest unit of evoluton.
There are $4^3$ possible triplets minus three stop-codons, resulting in a state space size of $61$ codons.
The standard assuption of independence is used, i.e. the substitutions at these three codon positions occur independently, thus only a single change per triplet can occur at a given time.
Selective pressure is the main force behind the molecular evolution.
Proteins are coded by a set of 20 amino acids, with each amino acid being coded by a codon triplet. 
Because there are 61 non-stop codons, inevitably some amino acids will be redundantly coded by more than 1 codon.
Such a change is called a synonymous substitution as they most likely will leave the protein function unchanged.
Conversly non-synonymous substitutions are more likely to affect the fitness of a particular organism.
Understanding and inferring the selective pressure is one of the central goals of molecular virology.
That is why most of the codon substitution models are parametrized in terms of the rate of non-synonymous (denoted by convention $\beta$) and synonymous ($\alpha$ by convention) substitutions.
Their ratio $\omega=\beta / \alpha$ is a standard measure of the selective pressure \citep{ThePhylogeneticHandbook}, which is sometimes also denoted $\omega = dN/dS$.
Prevalence of synonymous substitutions over non-synonymous ones is leading to a \emph{purifying (negative)} selection, and corresponds to the ratio $\omega <1$.
If non-synonymous substitutions accumulate at a faster rate than the synonymous substitutions this is the \emph{positive selection}, improving the fitness of the particular organism.
$\omega\approx1$ signifies neutral evolution.

There are several advantages in using codon models over nucleotide based substitution models.
Not all DNA positions evolve at the same rate, with non-synonymous substitutions occuring more frequently the synonymous substitutions.
Although this problem can be mitigated to some extent by using codon-positioned nucleotide substitution models, the fast evolving positions and the state space limited to 4 character alphabet leads to biased estimates of long evolutionary distances, as portrayed in the previous chapter (\ref{sub:deepRoot}).

@@

Different flavours of the GY94 model differ in the composition of the equilibrium codon frequency parameter $\pi_{j}$.
One approach is to model the codon frequencies as each having the same long-time frequency of appearing. 
Such model is reffered to as the GY94-Fequal.
In the GY94-F$1\times4$ model codon frequencies are expected from the four nucleotide frequencies, leading to 3 free parameters.
If the frequencies are parametrized according to three sets of nucleotide frequencies for the three codon positions, resulting in nine free parameters, the model is called the GY94-F$3\times4$.
Finally in the GY94-F61 model every codon triplet has it's own frequncy parameter, with all parameters summing to one, resulting in 60 free parameter that need to be estimated.

% pluses positive and negative of these models :)
